---
title: "Essential Statistics"
format: html
editor: visual
---

## Philosophical Foundations

* In scientific research we often deal with questions, problems, or claims.

* The main question being answered, problem being solved, or claim being verified or falsified should guide the decisions in research, including all decisions regarding data collection and analysis.

* So first we need to write down the specific research question, claim, or statement of the problem that our research is going to address.

* We need to have a theory of randomness: in what ways can random chance contributre to our conclusions?

* Summary statistics loses information regarding individuals. researchers often incorrectly then use them to infer information about individuals.

## Data

```{r Load-Package}
library(childesr)
library(tidyverse)
library(feather)
```

```{r Import-Data, eval=FALSE}
Providence_tokens <- get_tokens(collection = "Eng-NA", 
                                language = "eng", 
                                corpus = "Providence", 
                                token = "*")

Providence_utterances <- get_utterances(collection = "Eng-NA", 
                                        language = "eng", 
                                        corpus = "Providence")

saveRDS(Providence_utterances, "providence_utterances.rds")
```

```{r Load-Data}
Providence_utterances <- readRDS("providence_utterances.rds")
```

## Visualization and Descriptions

### Discrete

```{r}
Providence_utterances %>%
  ggplot(aes(x=target_child_sex)) + 
  geom_bar() +
  labs(x = "Target Child Sex", y = "Count") +
  scale_y_continuous(labels = scales::comma) +
  theme_classic()
```

```{r}
Providence_utterances %>%
  ggplot(aes(x=fct_infreq(speaker_role))) + 
  geom_bar() +
  labs(x = "Speaker Role", y = "Count") +
  theme_classic()
```

```{r}
Providence_utterances %>%
  ggplot(aes(x=type)) + 
  geom_bar() +
  labs(x = "Speaker Role", y = "Count") +
  theme_classic()
```

* Measure of Central Tendency: Mode -- The most frequently occurring value in a specific measurement.

```{r mode}
which.max(table(Providence_utterances$target_child_sex))
```

Relative Frequency

```{r}
Providence_utterances %>%
  count(target_child_sex) %>%
  mutate(relative_freq = n / sum(n),
         percent = relative_freq*100)
```


* Measure of Dispersion: Entropy

* Definition of Entropy: $$H(X) = -\sum_{i=1}^{n} p(x_i) \log_2 p(x_i)$$

```{r}
Providence_utterances %>%
  count(target_child_sex) %>%
  mutate(relative_freq = n / sum(n),
         entropy = -sum(relative_freq * log2(relative_freq)))
```

```{r}
Providence_utterances %>%
  count(speaker_role) %>%
  mutate(relative_freq = n / sum(n),
         entropy = -sum(relative_freq * log2(relative_freq)))
```


### Continuous

```{r}
Providence_utterances %>%
  ggplot(aes(x=target_child_age)) + 
  geom_histogram(binwidth = 1) +
  labs(x = "Target Child Age (months)", y = "Count") +
  theme_classic()
```

```{r boxplot}
ggplot(Providence_utterances, aes(target_child_sex, target_child_age)) +
  geom_boxplot() +
  labs(x = "Child Sex", y = "Child Age") +
  theme_classic()
```


```{r violin-plot}
ggplot(Providence_utterances, aes(target_child_sex, target_child_age)) +
  geom_violin() +
  labs(x = "Child Sex", y = "Child Age") +
  theme_classic()
```

```{r}
ggplot(Providence_utterances, aes(target_child_sex, num_tokens)) +
  geom_jitter(width = 0.1)
```

```{r}
Providence_utterances %>%
  group_by(target_child_age) %>%
  summarise(sum_tokens = sum(num_tokens))
```

```{r}
Providence_utterances %>%
  group_by(speaker_role, target_child_sex) %>%
  summarise(sum_tokens = sum(num_tokens)) %>%
  ggplot(aes(target_child_age, sum_tokens)) +
  geom_point() +
  labs(x = "Child Age", y = "Sum Tokens") +
  theme_classic()
```

```{r}
ggplot(Providence_utterances, aes(x=num_tokens, y=num_morphemes)) +
  geom_point() +
  theme_classic()
```



```{r}
cor(Providence_utterances$num_tokens, Providence_utterances$num_morphemes, use = "complete.obs")
```


```{r}
sum_tokens_by_age <-
  Providence_utterances %>%
  group_by(target_child_age) %>%
  summarise(sum_tokens = sum(num_tokens))
```

```{r}
ggplot(sum_tokens_by_age, aes(target_child_age, sum_tokens)) +
  geom_point() +
  labs(x = "Child Age", y = "Sum Tokens") +
  theme_classic()
```

## Models

* Models embody all your assumptions about what you are studying.

* What are the entities involved?

* How are they related to each other? What affects what? in what way? What causes what?

* Mathematics helps us be very precise about these.

* Regression is a method of estimating the parameters of your model.

```{r}
intercept_only_model <- lm(sum_tokens ~ 1 , data=sum_tokens_by_age)

summary(intercept_only_model)
```

```{r}
ggplot(sum_tokens_by_age, aes(x = target_child_age, y = sum_tokens)) +
  geom_point() +
  geom_hline(yintercept = coef(intercept_only_model)[1], color = "blue") + 
  theme_classic()
```

```{r}
intercept_val <- coef(intercept_only_model)[1]

ggplot(sum_tokens_by_age, aes(x = target_child_age, y = sum_tokens)) +
  geom_segment(aes(xend = target_child_age, yend = intercept_val), color = "red", alpha = 0.5) +
  geom_point() +
  geom_hline(yintercept = intercept_val, color = "blue")
```

```{r}
ggplot(mapping = aes(x = residuals(intercept_only_model))) +
  geom_histogram(bins = 30) +
  labs(x = "Residual", y = "Count") +
  theme_classic()
```



### Prediction and Error

* Mean Squared Error and Root Mean Squared Error

### Linear Regression


### Nonlinear Regression


## Distributions

```{r dist-setup}
library(ggplot2)
```

### Uniform Distribution

Every value between `a` and `b` is equally likely.

$$f(x) = \frac{1}{b - a} \quad \text{for } a \le x \le b$$

```{r uniform}
x <- seq(-0.5, 1.5, length.out = 200)
df <- data.frame(x, y = dunif(x, min = 0, max = 1))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Linear Distribution

A distribution whose PDF is a linear function on $[0, 1]$. For the density to be valid it must integrate to 1, requiring $\frac{a}{2} + b = 1$, and remain non-negative on the interval.

$$f(x) = ax + b \quad \text{for } 0 \le x \le 1, \quad \frac{a}{2} + b = 1$$

```{r linear}
a <- 1.6; b <- 1 - a / 2
x <- seq(0, 1, length.out = 200)
df <- data.frame(x, y = a * x + b)
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```


### Binomial Distribution

Models the number of successes in `n` independent trials, each with probability `p`.

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$$

```{r binomial}
x <- 0:10
df <- data.frame(x, y = dbinom(x, size = 10, prob = 0.5))
ggplot(df, aes(x, y)) + geom_col() + labs(x = "Successes", y = "Probability")
```

### Multinomial Distribution

Generalizes the binomial to `n` trials where each trial can result in one of `k` categories, each with probability $p_i$.

$$P(X_1 = x_1, \ldots, X_k = x_k) = \frac{n!}{x_1! \cdots x_k!} \, p_1^{x_1} \cdots p_k^{x_k}$$

```{r multinomial}
counts <- as.vector(rmultinom(1, size = 100, prob = c(0.2, 0.5, 0.3)))
df <- data.frame(category = c("A", "B", "C"), count = counts)
ggplot(df, aes(category, count)) +
  geom_col() + labs(x = "Category", y = "Count")
```

### Normal Distribution

Symmetric bell curve defined by a mean $\mu$ and standard deviation $\sigma$.

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$$

```{r normal}
x <- seq(-4, 4, length.out = 200)
df <- data.frame(x, y = dnorm(x))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Poisson Distribution

Models the count of events occurring at a constant average rate $\lambda$ over an interval.

$$P(X = k) = \frac{\lambda^k e^{-\lambda}}{k!}$$

```{r poisson}
x <- 0:15
df <- data.frame(x, y = dpois(x, lambda = 4))
ggplot(df, aes(x, y)) + geom_col() + labs(x = "Events", y = "Probability")
```

### Negative Binomial Distribution

Models the number of failures before the `r`-th success in a sequence of independent trials with success probability `p`.

$$P(X = k) = \binom{k + r - 1}{k}(1-p)^k p^r$$

```{r negative-binomial}
x <- 0:20
df <- data.frame(x, y = dnbinom(x, size = 5, prob = 0.4))
ggplot(df, aes(x, y)) + geom_col() + labs(x = "Failures", y = "Probability")
```

### Student's t Distribution

Used for inference about means when the sample size is small or the population variance is unknown. Heavier tails than the normal; approaches normal as $\nu \to \infty$.

$$f(x) = \frac{\Gamma\!\left(\frac{\nu+1}{2}\right)}{\sqrt{\nu\pi}\,\Gamma\!\left(\frac{\nu}{2}\right)} \left(1+\frac{x^2}{\nu}\right)^{-\frac{\nu+1}{2}}$$

```{r student-t}
x <- seq(-5, 5, length.out = 200)
df <- data.frame(x, y = dt(x, df = 5))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Log-Normal Distribution

A variable whose logarithm follows a normal distribution. Used to model positive-valued, right-skewed quantities.

$$f(x) = \frac{1}{x\sigma\sqrt{2\pi}}\, e^{-\frac{(\ln x - \mu)^2}{2\sigma^2}}$$

```{r log-normal}
x <- seq(0.01, 5, length.out = 200)
df <- data.frame(x, y = dlnorm(x, meanlog = 0, sdlog = 0.6))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Gamma Distribution

Generalizes the exponential distribution to model waiting times for multiple events. Parameterized by shape $\alpha$ and rate $\beta$.

$$f(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}\, x^{\alpha-1} e^{-\beta x}$$

```{r gamma}
x <- seq(0, 15, length.out = 200)
df <- data.frame(x, y = dgamma(x, shape = 3, rate = 0.5))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Beta Distribution

Models probabilities or proportions; defined on $[0, 1]$ and shaped by parameters $\alpha$ and $\beta$.

$$f(x) = \frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}$$

```{r beta}
x <- seq(0, 1, length.out = 200)
df <- data.frame(x, y = dbeta(x, shape1 = 2, shape2 = 5))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Dirichlet Distribution

Multivariate generalization of the Beta distribution. Models a vector of probabilities that sum to 1; parameterized by a concentration vector $\boldsymbol{\alpha}$.

$$f(x_1,\ldots,x_k) = \frac{1}{B(\boldsymbol{\alpha})}\prod_{i=1}^k x_i^{\alpha_i - 1}$$

```{r dirichlet}
alpha <- c(2, 5, 3)
g <- rgamma(length(alpha), shape = alpha)
probs <- g / sum(g)
df <- data.frame(category = c("A", "B", "C"), probability = probs)
ggplot(df, aes(category, probability)) +
  geom_col() + labs(x = "Category", y = "Probability")
```

### Chi-Squared Distribution

Sum of $k$ squared independent standard normal variables. Used in hypothesis testing and goodness-of-fit tests.

$$f(x) = \frac{x^{k/2-1} e^{-x/2}}{2^{k/2}\,\Gamma(k/2)}$$

```{r chi-squared}
x <- seq(0, 20, length.out = 200)
df <- data.frame(x, y = dchisq(x, df = 5))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Logistic Distribution

Symmetric and bell-shaped like the normal, but with heavier tails. Its CDF is the logistic function, making it central to logistic regression.

$$f(x) = \frac{e^{-(x-\mu)/s}}{s\left(1+e^{-(x-\mu)/s}\right)^2}$$

```{r logistic}
x <- seq(-6, 6, length.out = 200)
df <- data.frame(x, y = dlogis(x, location = 0, scale = 1))
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Value", y = "Density")
```

### Gompertz Distribution

Models survival and mortality rates; hazard increases exponentially over time. Parameterized by shape $\eta$ and rate $b$.

$$f(x) = \eta b\, e^{bx} \exp\!\left(-\eta\left(e^{bx} - 1\right)\right)$$

```{r gompertz}
x <- seq(0, 3, length.out = 200)
eta <- 1; b <- 2
y <- eta * b * exp(b * x) * exp(-eta * (exp(b * x) - 1))
df <- data.frame(x, y)
ggplot(df, aes(x, y)) + geom_line() + labs(x = "Time", y = "Density")
```


